---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Load Libraries

```{r}

# Load required libraries
library(tidyverse)
library(skimr)
library(janitor)
library(naniar)
library(ggplot2)
library(corrplot)
library(broom)
library(countrycode)

```

### Data Wrangling Functions

```{r}


load_energy_data <- function(url = "https://raw.githubusercontent.com/owid/energy-data/master/owid-energy-data.csv") {
  read_csv(url)
}

filter_energy_data <- function(df) {
  df %>%
    filter(!str_starts(iso_code, "OWID"), !is.na(country), !is.na(year)) %>%
    filter(year >= 2000) %>%
    select(where(~ sum(is.na(.)) < 0.4 * nrow(df))) %>%
    filter(!is.na(population), !is.na(gdp))
}

select_energy_columns <- function(df) {
  df %>%
    select(
      country, iso_code, year, population, gdp,
      electricity_generation,
      renewables_electricity, fossil_electricity,
      solar_electricity, wind_electricity, hydro_electricity,
      renewables_share_elec, coal_share_elec, gas_share_elec, oil_share_elec
    )
}

add_normalized_metrics <- function(df) {
  df %>%
    mutate(
      electricity_per_capita = electricity_generation / population,
      renewables_per_capita = renewables_electricity / population,
      fossil_per_capita = fossil_electricity / population,
      electricity_per_gdp = electricity_generation / gdp,
      gdp_per_electricity = gdp / (electricity_generation + 1)
    )
}

add_log_transforms <- function(df) {
  df %>%
    mutate(
      log_gdp = log(gdp + 1),
      log_population = log(population + 1),
      log_electricity = log(electricity_generation + 1)
    )
}

na_percentage <- function(df) {
  sapply(df, function(col) round(mean(is.na(col)) * 100, 2))
}

add_energy_ratios <- function(df) {
  df %>%
    mutate(
      fossil_to_renewable_ratio = fossil_electricity / (renewables_electricity + 1),
      fossil_share_elec = fossil_electricity / (electricity_generation + 1),
      solar_share = solar_electricity / (renewables_electricity + 1),
      wind_share = wind_electricity / (renewables_electricity + 1),
      hydro_share = hydro_electricity / (renewables_electricity + 1)
    )
}

add_classification_flags <- function(df) {
  df %>%
    mutate(
      high_renewable = if_else(renewables_share_elec > 50, 1, 0),
      transitioning = if_else(renewables_share_elec > fossil_share_elec, 1, 0)
    )
}

transform_energy_data <- function(df) {
  df %>%
    add_normalized_metrics() %>%
    add_log_transforms() %>%
    add_energy_ratios() %>%
    add_classification_flags()
}


```

### Load Dataset 2 - Electricity Generation Data

```{r}

# Load and process
energy_raw <- load_energy_data()
cat("Number of rows in Raw Dataset :",  nrow(energy_raw),'\n')
cat("Number of columns in Raw Dataset :", ncol(energy_raw),'\n')

energy_clean <- filter_energy_data(energy_raw)
cat("Number of rows in Filtered Dataset :",  nrow(energy_clean),'\n')
cat("Number of columns in Filtered Dataset :", ncol(energy_clean),'\n')

energy_selected <- select_energy_columns(energy_clean)

cat("Number of columns in Dataset after dropping unnecessary columns:", ncol(energy_selected),'\n')

energy_transformed <- transform_energy_data(energy_selected)

# Add continent info
energy_transformed <- energy_transformed %>%
  mutate(continent = countrycode(country, "country.name", "continent"))

cat("Number of columns in Dataset after creating new columns and transformations:", ncol(energy_transformed),'\n')


```

### Check for Duplicates and Missing Values

```{r}
### Checking for duplicated and % of missing values
energy_transformed %>% filter(duplicated(.))

### Checking for missing values
print("Percentage of Missing Values left :")
na_percentage(energy_transformed)
```

### Some latest stats about countries about renewable electricity

```{r, warning=FALSE}

# Filter for latest year
latest_year <- max(energy_transformed$year, na.rm = TRUE)
latest_data <- energy_transformed %>% filter(year == latest_year)

# Get top countries for different metrics
top_renewables <- latest_data %>% arrange(desc(renewables_electricity)) %>% slice(1)
top_generation <- latest_data %>% arrange(desc(electricity_generation)) %>% slice(1)
top_per_capita <- latest_data %>% arrange(desc(electricity_per_capita)) %>% slice(1)
top_gdp <- latest_data %>% arrange(desc(gdp)) %>% slice(1)
top_renew_share <- latest_data %>% arrange(desc(renewables_share_elec)) %>% slice(1)

# Print the results
cat(glue::glue("Year considered: {latest_year}\n"))
cat(glue::glue(" Country with most renewable electricity: {top_renewables$country} ({round(top_renewables$renewables_electricity, 2)} TWh)\n"))
cat(glue::glue(" Country with highest electricity generation: {top_generation$country} ({round(top_generation$electricity_generation, 2)} TWh)\n"))
cat(glue::glue(" Country with highest GDP: {top_gdp$country} (${format(round(top_gdp$gdp, 0), big.mark=',')})\n"))
cat(glue::glue(" Country with highest renewables share: {top_renew_share$country} ({round(top_renew_share$renewables_share_elec, 2)}%)\n"))
```

### Load CO2 and Clean Data - CO2 Emissions Dataset -1 

```{r}
library(tidyverse)
co2_data <- read_csv("https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv")

filter_co2_data <- function(df) {
  df %>%
    filter(!str_starts(iso_code, "OWID"), !is.na(country), !is.na(year)) %>%
    filter(year >= 2000) %>%
    select(where(~ sum(is.na(.)) < 0.4 * nrow(df))) %>%
    filter(!is.na(population), !is.na(gdp))
}

co2_data = filter_co2_data(co2_data)

co2_data = co2_data %>% select(-c(cement_co2,cement_co2_per_capita,co2_growth_abs,co2_growth_prct,co2_including_luc_growth_abs,co2_including_luc_growth_prct,cumulative_cement_co2,cumulative_co2_including_luc,cumulative_luc_co2,flaring_co2,flaring_co2_per_capita,cumulative_flaring_co2,share_global_flaring_co2,share_global_cumulative_flaring_co2))

#CO₂ emissions year-over-year change
co2_data = co2_data %>%
  group_by(country) %>%
  mutate(co2_pct_change = (co2 - lag(co2)) / lag(co2) * 100)
#Total CO₂ from fossil fuel types 
co2_data = co2_data %>%
  mutate(fossil_fuel_co2 = coal_co2 + oil_co2 + gas_co2)
#log gdp
co2_data = co2_data %>% mutate(log_gdp = log(gdp))
#log co2
co2_data = co2_data %>% mutate(log_co2 = log(co2))

```

### Merge Two Datasets to Single one - Using Country and Year as keys

```{r}
 

co2_transformed <- co2_data


co2_transformed <- co2_transformed %>%
  select(-iso_code, -population, -gdp, -log_gdp)
# 

energy_co2_merged <- left_join(
  energy_transformed,
  co2_transformed,
  by = c("country", "year")#,"iso_code","population","gdp")
)

#write_csv(energy_co2_merged, "/Users/manasamangipudi/Desktop/Semester-3/DataWrangling/Project/data/energy_co2_data_merged.csv")

```

```{r}
head(energy_co2_merged)
```

# Exploratory Data Analysis

## CO2 Dataset

### CO2 Emissions vs Population - How GDP impacts it

```{r}

library(ggrepel)
ggplot(energy_co2_merged %>% filter(year==2022), aes(x = population, y = co2, size = gdp, label = country)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text_repel(max.overlaps = 10, size = 5) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Fossil Fuel CO₂ Emissions vs Population (Bubble Size = GDP)",
    x = "Population (log scale)",
    y = "CO₂ Emissions (log scale)",
    size = "GDP"
  ) +
  theme_minimal()

```

### CO2 Emissions of Global Super Powers

```{r}


ggplot(energy_co2_merged %>% filter(country %in% c("United States","India","China","Russia","Japan")),aes(x=year,y=co2,color = country))+
  geom_line(size=1)+
  labs(title = "CO2 Emisions Over Time for Global Super Powers ",y="CO2 (Million Tonnes)")+
  theme_minimal()


```

### Paris Agreement Effect on CO2 Emissions 

```{r}

co2_change <- energy_co2_merged %>%
  filter(year %in% c(2015, 2022)) %>%
  select(country, year, co2,iso_code) %>%
  pivot_wider(names_from = year, values_from = co2, names_prefix = "co2_") %>%
  mutate(co2_diff_2015_2022 = co2_2022 - co2_2015)

co2_change %>%
  drop_na(co2_diff_2015_2022,iso_code) %>%
  arrange(co2_diff_2015_2022) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(country, co2_diff_2015_2022,decreasing = TRUE), y = co2_diff_2015_2022)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Countries with Most Decrease in CO₂ Emissions (2015–2022)",
    x = "Country",
    y = "Change in CO₂ Emissions (Mt)"
  ) +
  theme_minimal()

```

### CO2 Emissions vs GDP

```{r}


ggplot(energy_co2_merged, aes(x = gdp, y = co2)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "GDP vs CO2 Emissions")

```

## Electricity Generation Dataset : 

### Electricity Generation over Time by Continents

```{r}

ggplot(energy_co2_merged, aes(x = continent, y = electricity_per_capita)) +
  geom_boxplot(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Electricity per Capita by Continent")


energy_co2_merged %>%
  group_by(continent, year) %>%
  summarise(
    coal = sum(coal_share_elec, na.rm = TRUE),
    gas = sum(gas_share_elec, na.rm = TRUE),
    oil = sum(oil_share_elec, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c("coal", "gas","oil"), names_to = "source", values_to = "electricity") %>%
  ggplot(aes(x = year, y = electricity, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~continent, scales = "free_y") +
  theme_minimal() +
  labs(title = "Growth of Coal, Gas, Oil by Continent", y = "Electricity (TWh)")



energy_co2_merged %>%
  group_by(continent, year) %>%
  summarise(
    wind = sum(wind_share, na.rm = TRUE),
    solar = sum(solar_share, na.rm = TRUE),
    hydro = sum(hydro_share, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c("wind", "solar","hydro"), names_to = "source", values_to = "electricity") %>%
  ggplot(aes(x = year, y = electricity, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~continent, scales = "free_y") +
  theme_minimal() +
  labs(title = "Growth of Wind, Solar, Hydro by Continent", y = "Electricity (TWh)")

```
### Comparing Renewable and Fossil-fuel based Electricity Generation

```{r}


energy_by_continent <- energy_transformed %>%
  group_by(continent, year) %>%
  summarise(
    renewables = sum(renewables_electricity, na.rm = TRUE),
    fossil = sum(fossil_electricity, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c("renewables", "fossil"), names_to = "source", values_to = "electricity") %>%
  ungroup()

ggplot(energy_by_continent, aes(x = year, y = electricity, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~continent, scales = "free_y") +
  theme_minimal() +
  labs(title = "Renewable vs Non-Renewable Electricity Generation by Continent Over Time")

```

### Electricity vs GDP for year 2022 - Size representing Fossil_and_Renewable Ratio


```{r}


library(ggplot2)

ggplot(energy_co2_merged %>% filter(year==2022), aes(x = log_gdp, y = electricity_generation, size = fossil_to_renewable_ratio, label = country)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text_repel(max.overlaps = 10, size = 5) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  theme_minimal()

```

### Linear Relationship between GDP and Electricity Generation

```{r}


ggplot(energy_co2_merged, aes(y = gdp, x = electricity_generation)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "GDP vs Electricity Generation")

```

# Merged Dataset - EDA

### Global Trends of Electricity Generation and co2 emissions over time

```{r}

energy_co2_merged %>%
  group_by(year) %>%
  summarise(
    fossil = mean(fossil_electricity, na.rm = TRUE),
    co2 = mean(co2, na.rm = TRUE), 
    renewable = mean(renewables_electricity, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c("fossil", "co2", "renewable"), names_to = "source", values_to = "share") %>%
  ggplot(aes(x = year, y = share, color = source)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(
    title = "Global Trend of Fossil and co2 over time",
    x = "Year",
    y = "Share of Electricity (%)",
    color = "Source"
  )



```

### Continent-wise Trends of Electricity Generation and co2 emissions over time

```{r}
library(tidyverse)

energy_co2_merged %>%
  group_by(continent, year) %>%
  summarise(
    co2 = mean(co2, na.rm = TRUE),
    fossil = mean(fossil_electricity, na.rm = TRUE),
    renew = mean(renewables_share_elec, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  pivot_longer(cols = c(co2, fossil, renew), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = variable)) +
  geom_line(size = 1) +
  facet_wrap(~continent, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "CO₂ Emissions, Fossil and Renewable Electricity Trends by Continent",
    x = "Year",
    y = "Electricity Share/Emissions",
    color = "Metric"
  )

```

### Fossil Electricity and CO2 Emissions Linear Relationship

```{r}
energy_co2_merged %>%
  group_by(continent, year) %>%
  summarise(
    fossil_elec = mean(fossil_electricity, na.rm = TRUE),
    co2 = mean(co2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = fossil_elec, y = co2)) +
  geom_point(alpha = 0.6, color = "firebrick") +
  geom_smooth(method = "lm", color = "black") +
  facet_wrap(~continent, scales = "free") +
  labs(
    title = "Fossil Electricity vs CO₂ Emissions by Continent",
    x = "Fossil Electricity (TWh)",
    y = "CO₂ Emissions (Mt)"
  ) +
  theme_minimal()


```

### CO2 Emissions per unit of electricity generated over time

```{r}

energy_co2_merged %>%
  mutate(co2_per_twh = co2 /electricity_generation) %>%
  group_by(continent, year) %>%
  summarise(
    co2_efficiency = mean(co2_per_twh, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = co2_efficiency, color = continent)) +
  geom_line(size = 1.2) +
  labs(
    title = "CO₂ Emissions per Unit of Electricity Generated",
    y = "CO₂ / TWh",
    x = "Year"
  ) +
  theme_minimal()

```
### GHG Emissions per unit of electricity generated over time

```{r}


energy_co2_merged %>%
  mutate(ghg_per_twh = total_ghg /electricity_generation) %>%
  group_by(continent, year) %>%
  summarise(
    ghg_efficiency = mean(ghg_per_twh, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = ghg_efficiency, color = continent)) +
  geom_line(size = 1.2) +
  labs(
    title = "GHG Emissions per Unit of Electricity Generated",
    y = "GHG / TWh",
    x = "Year"
  ) +
  theme_minimal()

```

### Linear regression of CO2 against every other column sorted by estimate


```{r}

# 1. Filter for United States
us_data <- energy_co2_merged %>%
  filter(country == "United States") %>%
  select(-country, -iso_code, -continent, year)  # Remove non-numeric/grouping vars

# 2. Remove columns with all NA or zero variance
us_data <- us_data %>%
  select(where(is.numeric)) %>%
  select(where(~ sum(!is.na(.)) > 0)) %>%
  select(where(~ sd(., na.rm = TRUE) > 0))

# 3. Build models and extract p-values
results <- map_dfr(
  setdiff(names(us_data), "co2"),
  function(var) {
    df <- us_data %>% select(co2, !!sym(var)) %>% drop_na()
    if (nrow(df) < 10) return(NULL)  # skip if not enough data
    model <- lm(co2 ~ ., data = df)
    tidy(model) %>%
      filter(term != "(Intercept)") %>%
      mutate(variable = var)
  }
)

# 4. Print significant predictors (p >= 0.05)
significant <- results %>%
  filter(p.value < 0.05) %>%
  arrange(p.value)

head(significant %>% arrange(desc(estimate)))

```

### Fossil Fuel vs CO2 Emissions in 2022 (By Population)

```{r}


library(ggrepel)
ggplot(energy_co2_merged %>% filter(year==2022), aes(x = renewables_per_capita, y = co2, size = population, label = country)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text_repel(max.overlaps = 5, size = 4) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Fossil Fuel CO₂ Emissions vs Population (Bubble Size = Population)",
    x = "Renewables_per_capita",
    y = "CO₂ Emissions (log scale)",
    size = "Population"
  ) +
  theme_minimal()


```
